Starcode: an exact algorithm for sequence clustering

### 2014-07-01 doc/document.tex
Added lookup search section in the manuscript.

@@

\begin{algorithm}
  \caption{Starcode algorithm}
  \label{alg:starcode}
  \begin{algorithmic}[1]
    \State Define: $\tau$
    \State \textbf{Initialize variables:} $seed$, $start = 0$,
    $height$, $seq$, $trie$
    \State \textbf{Initialize containers:} $hits$, $pebbles$
    \State \textsc{read} sequence file
    \State $height \gets$ \textsc{determine} maximum sequence length
    \State \textsc{pad} sequences up to $height$
    \State \textsc{sort} sequences alphabetically
    \State $trie \gets$ \textsc{create} an empty trie of height $height$
    \State \textsc{insert} root node of $trie$ in $pebbles$ at depth 0
    \ForAll{sequences}
    \State $seq \gets$ \textsc{get} next sequence
    \State $seed \gets$ \textsc{length} of shared prefix
    between current and next $seq$
    \State \textsc{clear} $hits$
    \State \textsc{clear} $pebbles$ at depth $>start$
    \ForAll{$pebbles$ at depth $start$}
    \State $node \gets$ \textsc{get} next node from $pebbles$
    \State \textbf{call} \textsc{poucet}($node$, $seed$, $hits$)
    \EndFor
    \State \textsc{insert} $seq$ path in $trie$
    \State \textsc{process} $hits$ and \textsc{link} matches to $seq$
    \State $start \gets seed$
    \EndFor
  \end{algorithmic}
\end{algorithm}

\begin{algorithm}
  \caption{Poucet search algorithm}
  \label{alg:poucet}
  \begin{algorithmic}[1]
    \Procedure{poucet}{$node$, $seed$, $hits$}:
    \State \textsc{compute} $node$-specific column values following NW algorithm
    \ForAll{$child$ nodes in $node$}
    \State \textsc{compute} $child$-specific row values following NW algorithm
    \State \textsc{compute} center value using row and column
    values
    \If{center value $>\tau$} \Comment{Mismatches exceeded.}
    \State \textbf{continue} with next $child$
    \EndIf
    \If{$node$ depth = $height$} \Comment{Hit found.}
    \State \textsc{save} $node$ sequence in $hits$
    \State \textbf{continue} with next $child$
    \EndIf
    \If{$node$ depth $\leq seed$}
    \State \textsc{save} $node$ in pebbles at current depth
    \EndIf
    \State \textbf{call} \textsc{poucet}($child$, $seed$, $hits$)
    \EndFor
    \EndProcedure
  \end{algorithmic}
\end{algorithm}

>>>>

\subsection{Lookup search}
If the Levenshtein distance between two sequences of length $L$ is
less then or equal to $\tau$, then there must exist a perfect
match of length at least $k = L / (\tau+1)$, assuming for the time being
that $k$ is an integer.
We take advantage of this property to build a lookup table containing
all such $k$-mers from sequences inserted in the trie. When
a new sequence is queried, subsequences are first extracted from it
and queried against the lookup table. If none of the subsequences
are found, the minimum distance between the query and all the
sequences in the trie is at least $\tau+1$, and the trie-search can
therefore be omitted. This lookup is particulary useful for long query
sequences because in this case $k$ is large. The probability of an
exact match between subsequences of length $k$ is small and the
trie-search will be avoided for many such queries. The absence of match
is detected at the computational cost of several table lookups, compared
to the higher cost of measuring the Levenshtein distance between long
sequences.

In general $L / (\tau+1)$ is not an integer. To deal
with this complication, we divide the query into $\tau+2$ subsequences.
The first $k+1$ are called words and have target length
$(L-\tau) / (\tau+1)$. The last subsequence is called the tail and has
length $\tau$ for reasons explained below. The length of the words is
computed by integer division, and the remainder $R$ is distributed
equally among the last $R$ words. As a result, the words have a length
that differs by at most 1, with a total length equal to $L-\tau$.

A word at the beginning of the query is not a match for the same
word at the end of a sequence in the trie. For this reason, we query
the words against $k+1$ distinct lookup tables. In order to allow
insertions and deletions, the lookup search must also query shifted
versions of the words. In the worst case, only one word is found and the
mismatches are evenly distributed in the query sequence. The maximum
number of cumulative insertions/deletions at the $i$-th word
is $i-1$. Hence, the $i$-th word has to be shifted and queried against
the lookup table $2(i-1)+1$ times to cover all the possible cases.
The last word will therefore be queried $2\tau+1$ times, including
$\tau$ shifts on the right and $\tau$ shifts on the left. For the latter
we use the $\tau$ nucleotides stored in the tail.

@@

\begin{algorithm}
  \caption{Starcode algorithm}
  \label{alg:starcode}
  \begin{algorithmic}[1]
    \State \textbf{Define:} $\tau$
    \State \textbf{Variables:} $seed$, $start = 0$,
    $height$, $seq$, $trie$, $lastseq$, $k$
    \State \textbf{Containers:} $hits$, $pebbles$
    \State \textsc{read} sequence file
    \State $height \gets$ \textsc{determine} maximum sequence length
    \State \textsc{pad} sequences up to $height$
    \State \textsc{sort} sequences alphabetically
    \State $k \gets$ \textsc{compute} lookup word lengths
    \State $trie \gets$ \textsc{create} an empty trie of height $height$
    \State \textsc{insert} root node of $trie$ in $pebbles$ at depth 0
    \ForAll{sequences}
    \State $seq \gets$ \textsc{get} next sequence
    \If{at least one $k$-mer of $seq$ is in the lookup table}
    \State $seed \gets$ \textsc{length} of shared prefix
    between current and next sequence
    \State $start \gets$ \textsc{length} of shared prefix
    between $seq$ and $lastseq$
    \State \textsc{clear} hits
    \State \textsc{clear} $pebbles$ at depth $>start$
    \ForAll{$pebbles$ at depth $start$}
    \State $node \gets$ \textsc{get} next node from $pebbles$
    \State \textbf{call} \textsc{poucet}($seq$, $node$, $seed$, $hits$)
    \EndFor
    \State \textsc{process} $hits$ and \textsc{link} matches to $seq$
    \State $lastseq \gets seq$
    \EndIf
    \State \textsc{insert} $seq$ path in $trie$
    \State \textsc{insert} $seq$ $k$-mers into the lookup table
    \EndFor
  \end{algorithmic}
\end{algorithm}

\begin{algorithm}
  \caption{Poucet search algorithm}
  \label{alg:poucet}
  \begin{algorithmic}[1]
    \Procedure{poucet}{$query$, $node$, $seed$, $hits$}:
    \State \textsc{compute} $node$-specific column values following NW
    algorithm \Comment{Fig.1}
    \ForAll{$child$ nodes in $node$}
    \State \textsc{compute} $child$-specific row values following NW
    algorithm \Comment{Fig.1}
    \State \textsc{compute} center value using row and column values \Comment{Fig.1}
    \If{center value $>\tau$} \Comment{Mismatches exceeded.}
    \State \textbf{continue} with next $child$
    \EndIf
    \If{$node$ depth = $height$} \Comment{Hit found.}
    \State \textsc{save} $node$ sequence in $hits$
    \State \textbf{continue} with next $child$
    \EndIf
    \If{$node$ depth $\leq seed$}
    \State \textsc{save} $node$ in pebbles at current depth
    \EndIf
    \State \textbf{call} poucet($child$, $seed$, $hits$)
    \EndFor
    \EndProcedure
  \end{algorithmic}
\end{algorithm}
